Entropy-SGD: Biasing Gradient Descent Into Wide Valleys

### 2017-01-12 doc/rebuttal.txt

We thank the reviewers and the area chair for their insightful feedback. We have incorporated all comments into our current draft. We first discuss the updates to the paper that address common questions raised by the reviewers. We have posted comments to each reviewer to address specific questions.

Updates
-------

a) We have updated the experimental section of the paper with new experiments on MNIST, CIFAR-10 and two datasets on RNNs (PTB and char-LSTM). We also modified the algorithm to increase the value of the scoping parameter \gamma as training progresses instead of fixing it (Sec. 4.3). As a result of this, we can now train all our networks with only the local entropy loss instead of treating it as a regularizer (Eqn. 6).

We now plot the error curves in Figs. 4, 5 and 6 against the "effective number of epochs", i.e., the number of epochs of Entropy-SGD is multiplied by the number of Langevin iterations L (we set L=1 for SGD/Adam). Thus the x-axis is now a direct measure of the wall-clock time agnostic to the underlying implementation and proportional to the number of back-props as suggested, which makes the comparison fair.

@@

b) Improved exposition of the algorithm in Sec. 4.2 and discussion of experiments in Sec. 5.3, 5.4 that includes intuition for hyper-parameter tuning.

c) Appendix C discusses the possible connections to variational inference (this is the same as the discussion with the AC below). Sec. C.1 presents a comparison of local entropy and SGLD. We note here that our results using Entropy-SGD for both CNNs and RNNs are much better than SGLD is significantly smaller (~3x) wall-clock times.

Response to the reviewers:
-------------------------

a) smoothing of the original loss vs. local entropy
We discuss this in detail in the related work in Sec. 2 and Appendix C. While smoothing the original loss function using convolutions or averaging gradient over perturbations of weights reduces the ruggedness of the energy landscape, it does not help with sharp, narrow valleys. Local entropy introduces a measure that focuses on wide local minima in the energy landscape which is the primary reason for its efficacy. Smoothing the loss function can also, for instance, generate an artificial local minimum between two sharp valleys separated by a tall energy barrier which is detrimental to generalization.

b) unrealistic eigenvalue assumption in Sec. 4.3
We have clarified this point in Remark 4. Our analysis employs an assumption that the Hessian \nabla^2 f(x) does not have eigenvalues in the set [-2\gamma-c, c]. This is admittedly unrealistic, for instance, the eigenspectrum of the Hessian at a local minimum in Fig. 1 has a large fraction of its eigenvalues almost zero. Let us however remark that the bound on uniform stability in Thm. 3 by Hardt et al. assumes global conditions on the smoothness of the loss function; one imagines that Eqn. 9 remains qualitatively the same (with respect to the number of training iterations in particular) even if this assumption is violated to an extent before convergence happens. Obtaining a rigorous generalization bound without this assumption would require a dynamical analysis of SGD and seems out of reach of the current literature.

@@

------------------
AnonReviewer4:

>> how to reconcile "Entropy-SGD obtains better generalization but results in lower cross-entropy loss"
The confusion is probably caused by our omission, it should read "lower cross-entropy loss on the training set" (fixed now). This experimental observation suggests that there exist wide valleys that are deeper in the energy landscape that generalize well; (possibly) Entropy-SGD manages to find them while SGD gets stuck at a higher loss.

>> \rho = 0.01 on CIFAR-10 while \rho = 0 on MNIST
We have modified the algorithm to only train with local entropy (i.e., \rho = 0 always). Using scoping we can obtain better results both in terms of generalization error and wall-clock time.

>> discuss similarities to Hochreiter and Schmidhuber '97 [HS97]
We have modified our discussion of HS97 in Sec. 2 to include this discussion: While the motivations are exactly the same, the similarities with their formulation are only conceptual in nature, e.g., in a flat minimum, the local entropy is a direct measure of the width of the valley which is similar to their usage of Hessian. The Gibbs variant to averaging in weight space (Eqn. 33, pg. 22 of [HS97]) is similar to the averaging in Eqn. 7 of our paper.

We agree with the reviewer that the elaborate analysis of generalization of HS97 using the Gibbs formalism is a promising direction. In our case, we benefit from similar technical results introduced for uniform stability [BE02].

>> experiments are on a toy example
Please note that our results on CIFAR-10 are close to the best result on non-augmented CIFAR-10 (6.55% error using ELU units) and certainly the best reported results for the popular All-CNN network which is a medium-sized model with about 1.6 million weights. The largest model we experimented with is an LSTM on the PTB dataset with 66 million weights, please see the updates above for more details.

@@

[BE02] Olivier Bousquet and Andre Elisseeff. Stability and generalization. JMLR, 2002.

------------------
Csaba Szepesvari:

>> discussion of Baldassi et. al. '16 in related work
We have expanded the discussion of local entropy previously introduced by Baldassi et. al. in Sec. 2 and included due citations in Sec. 1 and 3 as well. This paper builds upon their work and generalizes it to models with continuous variables, deep networks in particular. After posting the first version of this paper, we have been collaborating with Baldassi et. al. for experiments on scoping techniques and they have co-authored this version of the paper.

@@

>> Eqn. 8 should have f(x')
Thanks, it is fixed now.

>> experiments on RNNs
Please see updates to the paper above (a), we have included results for word-level and character-level text prediction on PTB and char-LSTM with War and Peace. We not only obtain a better test perplexity than a competitive baseline with SGD, but also train in half as much wall-clock time.

------------------
AnonReviewer2:

>> Thm. 3 does not have the constant c in Hardt et. al. '15
We set the constant c of Hardt et. al., to 1 in our statement to avoid confusion with our constant, with is also called c (the learning \eta_t \leq 1/t instead of the original \eta_t \leq c/t). This does not change the result qualitatively.

### 2017-01-12 doc/chaudhari.choromanska.ea.iclr17.tex
wip on rebuttal.txt

\author{Pratik Chaudhari$^{1}$, Anna Choromanska$^{2}$, Stefano Soatto$^{1}$, Yann LeCun$^{2,3}$, Carlo Baldassi$^{4}$,\\[0.03in]
\textbf{Christian Borgs$^{5}$, Jennifer Chayes$^{5}$, 
Riccardo Zecchina$^{4}$}\\[0.05in]
$^{1}$ Computer Science, University of California, Los Angeles.\\
$^{2}$ Courant Institute of Mathematical Sciences, New York University.\\
$^{3}$ Facebook AI Research, New York.\\
$^{4}$ Politecnico di Torino.\\
$^{5}$ Microsoft Research New England, Cambridge.\\ [0.05in]
{\footnotesize
Email:\ \href{mailto:pratikac@ucla.edu}{pratikac@ucla.edu},
\href{mailto:achoroma@cims.nyu.edu}{achoroma@cims.nyu.edu},
\href{mailto:soatto@cs.ucla.edu}{soatto@cs.ucla.edu},
\href{mailto:yann@cs.nyu.edu}{yann@cs.nyu.edu},
\href{mailto:carlobaldassi@gmail.com}{carlobaldassi@gmail.com}}\\[0.03in]
{\footnotesize
\hspace{0.33in} \href{mailto:borgs@microsoft.com}{borgs@microsoft.com},
\href{mailto:jchayes@microsoft.com}{jchayes@microsoft.com},
\href{mailto:riccardo.zecchina@polito.it}{riccardo.zecchina@polito.it}
}}

@@

Motivated by the same final goal, viz., flat local minima, the authors in~\citet{hochreiter1997flat} introduce hard constraints on the training loss and the width of local minima and show using the Gibbs formalism~\citep{haussler1997mutual} that this leads to improved generalization. As the authors discuss, the effect of hyper-parameters for the constraints is intricately tied together and they are difficult to choose even for small problems. Our local entropy based objective naturally balances the energetic term (training loss) and the entropic term (width of the valley) and also results in a more scalable algorithm. The ``scope'' $\g$ affords more control on the optimization process than hard constraints and as we discuss in Sec.~\ref{s:scoping}, this also provides significant computational advantages. The Gibbs formalism used in the analysis of~\citet{hochreiter1997flat} is however a very promising direction to understanding generalization in deep networks in the context of this paper.
%, it bears supercial similarities to the uniform stability analysis~\citep{haussler1997mutual} we employ.
%Recent work by~\citet{keskar2016large} is relevant in the context of the main theme of our paper. The authors estimate the loss in a neighborhood of the weights to argue that small batch size in SGD (i.e., larger gradient noise) generalizes better than large mini-batches and also results in significantly flatter minima.

>>>>

\author{Pratik Chaudhari$^{1}$, Anna Choromanska$^{2}$, Stefano Soatto$^{1}$, Yann LeCun$^{2,3}$, Carlo Baldassi$^{4}$,\\[0.03in]
\textbf{Christian Borgs$^{5}$, Jennifer Chayes$^{5}$, 
Riccardo Zecchina$^{4}$}\\[0.05in]
$^{1}$ Computer Science, University of California, Los Angeles.\\
$^{2}$ Courant Institute of Mathematical Sciences, New York University.\\
$^{3}$ Facebook AI Research, New York.\\
$^{4}$ Politecnico di Torino.\\
$^{5}$ Microsoft Research New England, Cambridge.\\ [0.05in]
{\footnotesize
Email:\ \href{mailto:pratikac@ucla.edu}{pratikac@ucla.edu},
\href{mailto:achoroma@cims.nyu.edu}{achoroma@cims.nyu.edu},
\href{mailto:soatto@cs.ucla.edu}{soatto@cs.ucla.edu},
\href{mailto:yann@cs.nyu.edu}{yann@cs.nyu.edu},
\href{mailto:carlo.baldassi@polito.it}{carlo.baldassi@polito.it},}\\[0.03in]
{\footnotesize
\hspace{0.33in} \href{mailto:borgs@microsoft.com}{borgs@microsoft.com},
\href{mailto:jchayes@microsoft.com}{jchayes@microsoft.com},
\href{mailto:riccardo.zecchina@polito.it}{riccardo.zecchina@polito.it}
}}

@@

\pc{Motivated by the same final goal, viz., flat local minima, the authors in~\citet{hochreiter1997flat} introduce hard constraints on the training loss and the width of local minima and show using the Gibbs formalism~\citep{haussler1997mutual} that this leads to improved generalization. As the authors discuss, the effect of hyper-parameters for the constraints is intricately tied together and they are difficult to choose even for small problems. Our local entropy based objective naturally balances the energetic term (training loss) and the entropic term (width of the valley) and also results in a more scalable algorithm. The ``scope'' $\g$ affords more control on the optimization process than hard constraints and as we discuss in Sec.~\ref{s:scoping}, this also provides significant computational advantages. The Gibbs formalism used in the analysis of~\citet{hochreiter1997flat} is however a very promising direction to understanding generalization in deep networks in the context of this paper.}{While the motivations are exactly the same, the similarities with their formulation are only conceptual in nature, e.g., in a flat minimum, the local entropy is a direct measure of the width of the valley which is similar to their usage of Hessian. The Gibbs variant to averaging in weight space (Eqn. 33, pg. 22 of [HS97]) is similar to the averaging in Eqn. 7 of our paper.}
